Commutative Character Sheaves

### 2015-10-13 followup.tex
Fixed the definition of H-equivariant character sheaf.
I was missing the compatibility of nu and mu.

\begin{abstract}
We introduce and establish the basic properties of  the category of \emph{commutative character sheaves} on smooth group schemes $G$ over finite fields $\Fq$. 
We use this category to geometrize all generic types for supercuspidal representations of $p$-adic groups. 
\end{abstract}

@@

\begin{definition}\label{def:CCS}
A linear character sheaf on $G$ is a \emph{commutative character sheaf} if there is an isomorphism $m^* \cs{L} \to m\op^* \cs{L}$ of local systems on $G\times G$ such that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=30]
   m^*\cs{L} \arrow{d} \arrow{r}{\mu} & \cs{L}\boxtimes\cs{L} \arrow{d}{\vartheta}\\
   m\op^*\cs{L} \arrow{r}{{\theta}^*\mu} &  {\bar \theta}^*(\cs{L}\boxtimes\cs{L})
  \end{tikzcd}
  \]
We write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

\begin{lemma}
If the isomorphism $m^* \cs{L} \to m\op^* \cs{L}$ exists and satisfies the condition of Definition~\ref{def:CCS}, then it is unique.
\end{lemma}

@@

Let $G$ be a smooth group scheme.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the smooth morphism $G\times G \to G$ given by $(x,y)\mapsto xyx^{-1}y^{-1}$. Then $G \times G \to G\der$ is a dominant morphism \cite{SGA3}*{{VI}$_B$, 7.2}
and $G$ is commutative if and only if $G\der$ is trivial.

While the subgroup $G\der$ is smooth, it need not be connected. The restriction of $G\times G\to G\der$ to identity components gives a dominant morphism $G^0\times G^0 \to G^0\der$ by [SGA3, Tome 6B, Expos\'e 7, Prop 7.8]

The quotient $G\ab\ceq G/G\der$ exists in schemes over $\Fq$ by SGA3, Expos\'e VI, Tome B, Th\'eor\`eme 3.2, which also shows that the quotient $G \to G\ab$ is a faithfully flat and a $G$-torsor in the fppf topology.
\[
\begin{tikzcd}
{} & G\times G \arrow{d}{} &&&\\
1 \arrow{r} & G\der \arrow{r}{} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]

@@

Another hypothesis: suppose the smooth morphism $G\times G\to G\der$ is of relative dimension on each connected component of its image in $G\der$.\todo{This is another condition on $G$. Does it come from smoothness of $G$?}

\begin{theorem}\label{thm:Gab}
Under these hypotheses, pull-back along the abelianization $q : G \to G\ab$
\[
q^* : \CCS(G\ab) \to \CCS(G)
\]
is an equivalence of categories.
\end{theorem}

@@

Let $G$ and $H$ be group schemes; let $n : H \times G\to G$ be a group action and write $p : H\times G \to G$ for projection. 
An $H$-equivariant Weil local sytem on $G$ is a Weil local system $\cs{L}$ on $G$ together with an isomorphism  $\nu : n^*\cs{L} \to p^*\cs{L}$ of Weil local systems on $G\times G$ such that the following diagram commutes,
\[
\begin{tikzcd}
\ &  n_0^*  n^*\, \mathcal{L} \arrow{r}{n_0^*\nu} & n_0^* p^*\, \mathcal{L} & \\
n_1^* n^*\, \mathcal{L} \arrow{dr}[swap]{n_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_2^* n^*\, \mathcal{L} \\
& n_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_2^* p^*\, \mathcal{L}, & 
\end{tikzcd}
\]
where $n_0(a,b,x) = (a,n(b,x))$ and $n_1(a,b,x) = (n(ab),x)$ and $p_2(a,b,x) = (b,x)$.
Morphisms of $H$-equivariant Weil local systems are morphisms of Weil local systems $\alpha: \cs{L}\to \cs{L}'$ for which . . .

An $H$-equivariant linear character sheaf on $G$ is a triple $(\cs{L},\mu, \nu)$ where $(\cs{L},\mu)$ with no condition tying $\mu$ and $\nu$ together.
Write $\CS_H(G)$ for the category of $H$-equivariant linear character sheaves on $G$.\todo{Under construction}

Finally, an $H$-equivariant commutative character sheaf on $G$ is an $H$-equivariant linear character sheaf $(\cs{L},\mu, \nu)$ for whuch

\begin{lemma}\label{lem:equivariant}
Categories $\CCS(G)$ and $\CCS_G(G)$ are equivalent.
\end{lemma}

\begin{proof}
Let $a : G \times G \to G$ be the conjugation action of $G$ on $G$ and let $p : G \times G \to G$ be projection to the second factor.
Then $a = m \circ (m \times (\iota \circ p_1))$ while $m\op \circ (m \times (\iota \circ p_1)) = p$;
likewise, $m = a \circ (p_1\times m\op)$ while $p \circ (p_1\times m\op) = m\op$.

Let $(\cs{L},\mu)$ be a commutative character sheaf on $G$.
Let $c: m^*\cs{L} \to m\op^*\cs{L}$ be an isomorphism satisfying the condition of Definition~\ref{def:CCS}. Then
\begin{align*}
a^* \cs{L} 
&= (m \circ (m \times (\iota \circ p_1)))^* \cs{L}\\
&= (m \times (\iota \circ p_1))^* m^* \cs{L}\\
&\iso (m \times (\iota \circ p_1))^* m\op^* \cs{L}\\
&= (m\op \circ (m \times (\iota \circ p_1)))^* \cs{L}\\
&= p^*\cs{L}.
\end{align*}
The isomorphism above is 
\[
(m \times (\iota \circ p_1))^* c: a^*\cs{L} \to p^*\cs{L},
\]
 so $(\cs{L}, (m \times (\iota \circ p_1))^* c)$ is a $G$-equivariant commutative character sheaf.

Let $(\cs{L}, \nu)$ be an equivariant linear character sheaf.
Then
\begin{align*}
m^* \cs{L} 
&= (a \circ (p_1 \times m\op))^* \cs{L}\\
&= (p_1 \times m\op)^* a^* \cs{L}\\
&\iso (p_1 \times m\op)^* p^* \cs{L}\\
&= (p\circ (p_1 \times m\op))^* \cs{L}\\
&= m\op^* \cs{L}.
\end{align*}
The isomorphism above is
\[
(p_1 \times m\op)^* \nu : m^* \cs{L}  \to m\op^* \cs{L},
\]
so $(\cs{L},\mu)$ is a commutative character sheaf.
\todo{This proof is under construction.}
\end{proof}

@@

Write $n : H \times_S G \to G$ for the left action of $H$ on $G$; let $p : H\times_S G \to G$ be projection.

We now suppose that $H$ is normal in $G$ and that the fppf-quotient $G \to G'\ceq H\backslash G$ exists in group schemes over $\Fq$.
Then $q : G \to G'$ is a regular epimorphism with kernel pair $(n,p)$ in the category of schemes.
\[
\begin{tikzcd}
H\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G'
\end{tikzcd}
\]
%In particular, $G \times_{G'} G \iso H\times_S G'$, so $G \to G'$ is a $G$-torsor in the fppf topology.
In this case we may factor the functor $\CCS(G') \to \CCS(G)$  . . . 
\[
\begin{tikzcd}
\CCS(G) &&\arrow{ll}[swap]{q^*} 
\CCS(G') \arrow{dl}{\text{`comparison'}}\\
& \arrow{ul}{\text{forget}} \CCS_{H}(G) & 
\end{tikzcd}
\]

@@

%Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Let $n : G\der \times G \to G$ be the restriction of $m: G\times G \to G$.
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L}.
\end{align*}
Thus, every commutative character sheaf $\cs{L}$ on $G$ is $G\der$-equivariant.
It now follows from Lemma~\ref{lem:Lder} that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence. 
\end{proof}

>>>>

\begin{abstract}
We introduce and establish the basic properties of  the category of \emph{commutative character sheaves} on smooth group schemes $G$ over finite fields $\Fq$. 
We use this category to geometrize types for supercuspidal representations of $p$-adic groups. 
\end{abstract}

@@

\begin{definition}\label{def:CCS}
A linear character sheaf on $G$ is a \emph{commutative character sheaf} if there is an isomorphism $\xi: m^* \cs{L} \to m\op^* \cs{L}$ of [linear character sheaves?] on $G\times G$ such that the following diagram commutes.
  \[
  \begin{tikzcd}[row sep=30]
   m^*\cs{L} \arrow{d}{\xi} \arrow{r}{\mu} & \cs{L}\boxtimes\cs{L} \arrow{d}{\vartheta}\\
   m\op^*\cs{L} \arrow{r}{{\theta}^*\mu} &  {\bar \theta}^*(\cs{L}\boxtimes\cs{L})
  \end{tikzcd}
  \]
We write $\CCS(G)$ for the full subcategory of $\CS(G)$ consisting of commutative character sheaves.
 \end{definition}

We will see that this condition is not trivial: even on commutative group schemes over $\Fq$, not every linear character sheaf is commutative; even on connected reductive algebraic groups over $\Fq$, not every linear character sheaf is commutative.

\begin{lemma}\label{lem:symmetry}
If the isomorphism $\xi: m^* \cs{L} \to m\op^* \cs{L}$ exists and satisfies the condition of Definition~\ref{def:CCS}, then it is unique.
\end{lemma}

@@

Let $G$ be a smooth group scheme.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the smooth morphism $G\times G \to G$ given by $(x,y)\mapsto xyx^{-1}y^{-1}$. Then $G \times G \to G\der$ is a dominant morphism \cite{SGA3}*{{VI}$_B$, 7.2}.
Moreover, $G$ is commutative if and only if $G\der$ is trivial.
%
While the subgroup $G\der$ is smooth, it need not be connected. 
The restriction of $G\times G\to G\der$ to identity components gives a dominant morphism $G^0\times G^0 \to G^0\der$ by [SGA3, Tome 6B, Expos\'e 7, Prop 7.8]

The quotient $G\ab\ceq G/G\der$ exists in schemes over $\Fq$ by SGA3, Expos\'e VI, Tome A, Th\'eor\`eme 3.2, which also shows that the quotient $G \to G\ab$ is a faithfully flat and a $G\der$-torsor in the fppf topology.
\[
\begin{tikzcd}
{} & G\times G \arrow{d}{} &&&\\
1 \arrow{r} & G\der \arrow{r}{} & G \arrow{r}{q} & G\ab \arrow{r} & 1
\end{tikzcd}
\]

@@

Provisional hypothesis: suppose the smooth morphism $G\times G\to G\der$ is of relative dimension on each connected component of its image in $G\der$...\todo{This is another condition on $G$. Does it come from smoothness of $G$?}

\begin{theorem}\label{thm:Gab}
Under this hypothesis, pull-back along the abelianization $q : G \to G\ab$
\[
q^* : \CCS(G\ab) \to \CCS(G)
\]
is an equivalence of categories.
\end{theorem}

@@

Let $G$ and $H$ be group schemes; let $a : H \times G\to G$ be a group action and write $p : H\times G \to G$ for projection. 
An $H$-equivariant Weil local sytem on $G$ is a Weil local system $\cs{L}$ on $G$ together with an isomorphism  $\nu : a^*\cs{L} \to p^*\cs{L}$ of Weil local systems on $G\times G$ such that the following diagram of local systems on $H\times H\times G$ commutes,
\[
\begin{tikzcd}
\ &  a_0^*  a^*\, \mathcal{L} \arrow{r}{a_0^*\nu} & a_0^* p^*\, \mathcal{L} & \\
a_1^* a^*\, \mathcal{L} \arrow{dr}[swap]{a_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_3^* a^*\, \mathcal{L} \\
& a_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_3^* p^*\, \mathcal{L}, & 
\end{tikzcd}
\]
where $a_0(x,y,z) = (x,a(y,z))$ and $a_1(x,y,z) = (a(x,y),z)$ and $p_3(x,y,z) = (y,z)$.
We remark that, on stalks, this condition is ...
\[
\begin{tikzcd}
\ &  \cs{L}_{x\cdot(y\cdot z)} \arrow{r}{\nu_{x,y\cdot z}} & \cs{L}_{y\cdot z} & \\
\cs{L}_{(x\cdot y)\cdot z} \arrow{dr}[swap]{\nu_{x\cdot y,z}} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{\nu_{y,z}} \cs{L}_{y\cdot z} \\
& \cs{L}_z \arrow[equal]{r} & \cs{L}_z & 
\end{tikzcd}
\]

Morphisms of $H$-equivariant Weil local systems are morphisms of Weil local systems $\alpha: \cs{L}\to \cs{L}'$ for which . . .

An $H$-equivariant linear character sheaf on $G$ is a triple $(\cs{L},\mu, \nu)$ where $(\cs{L},\mu)$ is a linear character sheaf and $\nu$ is compatible with $\mu$ in the sense that the following diagram of local systems on $H \times G\times G$ commutes ....
On stalks, this condition is
\[
\begin{tikzcd}
\ &  \cs{L}_{h\cdot (xy)} \arrow{r}{} & \cs{L}_{xy} & \\
\cs{L}_{(h\cdot x)(h\cdot y)} \arrow{dr}[swap]{} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{} \cs{L}_{xy} \\
& \cs{L}_{h\cdot x}\otimes \cs{L}_{h\cdot y} \arrow{r} & \cs{L}_x\otimes \cs{L}_y & 
\end{tikzcd}
\]

Write $\CS_H(G)$ for the category of $H$-equivariant linear character sheaves on $G$.
%\todo{Under construction}
%
Finally, an $H$-equivariant commutative character sheaf on $G$ is an $H$-equivariant linear character sheaf $(\cs{L},\mu, \nu)$ for whuch

\begin{lemma}
If $(\cs{L},\mu,\nu)$ is an equivariant linear character sheaf for the conjugation action of $G$ on $G$, then $(\cs{L},\mu)$ is a commutative character sheaf.
\end{lemma}

\begin{proof}
Let $a : G \times G \to G$ be the conjugation action $(g,h) \mapsto ghg^{-1}$ and let $p : G \times G \to G$ be projection to the second factor.
Let $\iota : G\to G$ be inversion for the group operation $m : G\times G\to G$.
Then $m = a \circ (p_1\times m\op)$ while $p \circ (p_1\times m\op) = m\op$.
Now, let $(\cs{L}, \mu, \nu)$ be an equivariant linear character sheaf.
Then
\begin{align*}
m^* \cs{L} 
&= (a \circ (p_1 \times m\op))^* \cs{L}\\
&= (p_1 \times m\op)^* a^* \cs{L}\\
&\iso (p_1 \times m\op)^* p^* \cs{L}\\
&= (p\circ (p_1 \times m\op))^* \cs{L}\\
&= m\op^* \cs{L}.
\end{align*}
The isomorphism above is
\[
\xi \ceq (p_1 \times m\op)^* \nu : m^* \cs{L}  \to m\op^* \cs{L}
\]
Observe that $p_1 \times m\op : G\times G \to G\times G$ is an isomorphism of varieties . . . 
The isomorphism $\xi$ satisfies the condition of Definition~\ref{def:CCS}. \todo{Checking this.}

Thus, $(\cs{L},\mu)$ is a commutative character sheaf.
\end{proof}

\begin{remark}
By Lemma~\ref{lem:symmetry}, the isomorphism $\xi_\nu$ is uniquely determined by $\mu$.
%This is the first hint that the forgetful functor may be an equivalence.
\end{remark}

\begin{lemma}\label{lem:equivariant}
The forgetful functor $\CCS_G(G)\to \CCS(G)$, given by $(\cs{L},\mu,\nu) \mapsto (\cs{L},\mu)$, is an equivalence.
\end{lemma}

\begin{proof}
Now, let $(\cs{L},\mu)$ be a commutative character sheaf.
Let $\xi: m^*\cs{L} \to m\op^*\cs{L}$ be an isomorphism satisfying the condition of Definition~\ref{def:CCS}.
Observe that $a = m \circ (m \times (\iota \circ p_1))$ while $m\op \circ (m \times (\iota \circ p_1)) = p$.
Then
\begin{align*}
a^* \cs{L} 
&= (m \circ (m \times (\iota \circ p_1)))^* \cs{L}\\
&= (m \times (\iota \circ p_1))^* m^* \cs{L}\\
&\iso (m \times (\iota \circ p_1))^* m\op^* \cs{L}\\
&= (m\op \circ (m \times (\iota \circ p_1)))^* \cs{L}\\
&= p^*\cs{L}.
\end{align*}
The isomorphism above is 
\[
\nu_\xi \ceq (m \times (\iota \circ p_1))^* \xi: a^*\cs{L} \to p^*\cs{L},
\]
 so $(\cs{L}, \mu, \nu_\xi)$ is a $G$-equivariant commutative character sheaf.

 forgetful  rule $(\cs{L},\mu)\to (\cs{L},\mu,\nu_\xi)$ and 
To define the promised adjunction, I need to modify the functors slighlty . . . 
\todo{This proof is under construction.}
\end{proof}

@@

We now suppose that $H$ is normal in $G$ and that the fppf-quotient $G \to G'\ceq H\backslash G$ exists in group schemes over $\Fq$.
Then $q : G \to G'$ is a regular epimorphism with kernel pair $(n,p)$ in the category of schemes.
\[
\begin{tikzcd}
H\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G'
\end{tikzcd}
\]

Let $j : H \hookrightarrow G$ be the kernel of $q : G \to G'$ and let $m : G\times G \to G$ be multiplication for $G$. 
Write $n : H \times_S G \to G$ for the left action of $H$ on $G$; this is obtained by restricting the action $m : G\times G\to G$ to the subgroup $H\times G$. 
Let $n : H \times G \to G$ be the restriction of $m: G\times G \to G$.
Then $n = m \circ (j \times \id)$, so

 let $p : H\times_S G \to G$ be projection.

%In particular, $G \times_{G'} G \iso H\times_S G'$, so $G \to G'$ is a $G$-torsor in the fppf topology.
In this case we may factor the functor $\CCS(G') \to \CCS(G)$  . . . 
\[
\begin{tikzcd}
\CCS(G) &&\arrow{ll}[swap]{q^*} 
\CCS(G') \arrow{dl}{\text{`comparison'}}\\
& \arrow{ul}{\text{forget}} \CCS_{H}(G) & 
\end{tikzcd}
\]

@@

%Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Let $n : H \times G \to G$ be the restriction of $m: G\times G \to G$.
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L};
\end{align*}
the isomorphism above is 
\[
\nu_\mu \ceq(j \times \id))^* \mu: n^*\cs{L} \to p^*\cs{L}.
\]
Thus, every commutative character sheaf $\cs{L}$ on $G$ is $H$-equivariant.
It now follows from Lemma~\ref{lem:Lder} that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence. 
\end{proof}
